---
title: "Efficient genome searching with Biostrings and the BSgenome data packages"
author: "Hervé Pagès"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
package: BSgenome
vignette: >
  %\VignetteIndexEntry{Efficient genome searching with Biostrings and the BSgenome data packages}
  %\VignetteKeywords{BSgenome, genome, DNA, RNA, Sequence, Biostrings, Sequence alignment, SNPs}
  %\VignettePackage{BSgenome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
library(BiocStyle)
```

# The Biostrings-based genome data packages

The Bioconductor project provides data packages that contain the full genome
sequences of a given organism. These packages are called
*Biostrings-based genome data packages* because the sequences they
contain are stored in some of the basic containers defined in the
`r Biocpkg('Biostrings')` package, like the `DNAString`,
the `DNAStringSet` or the `MaskedDNAString` containers.
Regardless of the particular sequence data that they contain, all the
Biostrings-based genome data packages are very similar and can be
manipulated in a consistent and easy way.
They all require the `r Biocpkg('BSgenome')` package in order to work properly.
This package, unlike the Biostrings-based genome data packages,
is a software package that provides the infrastructure needed to
support them (this is why the Biostrings-based genome data packages
are also called *BSgenome data packages*).
The `r Biocpkg('BSgenome')` package itself requires the `r Biocpkg('Biostrings')`
package.

See the man page for the `available.genomes` function
(`?available.genomes`) for more information about how to get
the list of all the BSgenome data packages currently available in your
version of Bioconductor (you need an internet connection so that
`available.genomes` can query the Bioconductor package
repositories).

Note that the **BSgenomeForge** package provides tools that you
can use to make your own BSgenome data package.

# Finding an arbitrary nucleotide pattern in a chromosome

In this section we show how to find (or just count) the occurences of some
arbitrary nucleotide pattern in a chromosome. The basic tool for this is
the `matchPattern` (or `countPattern`) function
from the `r Biocpkg('Biostrings')` package.

First we need to install and load the BSgenome data package for the
organism that we want to look at. In our case, we want to search
chromosome I of *Caenorhabditis elegans*.

UCSC provides several versions of the C. elegans genome: ce1, ce2 and ce4.
These versions correspond to different *releases* from WormBase,
which are the WS100, WS120 and WS170 releases, respectively.
See <http://genome.ucsc.edu/FAQ/FAQreleases#release1> for the list
of all UCSC genome releases and for the correspondance between UCSC
versions and release names.

The BSgenome data package for the ce2 genome is
**BSgenome.Celegans.UCSC.ce2**. Note that ce1 and ce4 are not
available in Bioconductor but they could be added if there is demand for
them.

See `?available.genomes` for how to install
**BSgenome.Celegans.UCSC.ce2**.
Then load the package and display the single object defined in it:

```{r b1}
library(BSgenome.Celegans.UCSC.ce2)
ls("package:BSgenome.Celegans.UCSC.ce2")
genome <- BSgenome.Celegans.UCSC.ce2
genome
```

`genome` is a `BSgenome` object:

```{r b2}
class(genome)
```

When displayed, some basic information about the origin of the
genome is shown (organism, genome, provider, etc...)
followed by the index of *single* sequences and eventually
an additional index of *multiple* sequences.
Methods (adequately called *accessor methods*) are defined
for individual access to this information:

```{r b3}
metadata(genome)
seqnames(genome)
seqinfo(genome)
```

See the man page for the `BSgenome` class (`?BSgenome`)
for a complete list of accessor methods and their descriptions.

Now we are ready to display chromosome I:

```{r b4}
genome$chrI
```

Note that this chrI sequence corresponds to the *forward* strand
(aka *direct* or *sense* or *positive* or *plus* strand)
of chromosome I.
UCSC, and genome providers in general, don't provide files containing the
nucleotide sequence of the *reverse* strand (aka *indirect*
or *antisense* or *negative* or *minus* or *opposite* strand)
of the chromosomes because these sequences can be deduced from the *forward*
sequences by taking their reverse complements.
The BSgenome data packages are no exceptions: they only
provide the *forward* strand sequence of every chromosome.
See `?reverseComplement` for more details about the reverse
complement of a `DNAString` object.
It is important to remember that, in practice, the *reverse* strand
sequence is almost never needed.
The reason is that, in fact, a *reverse* strand analysis
can (and should) always be transposed into a *forward* strand analysis.
Therefore trying to compute the *reverse* strand sequence of an entire
chromosome by applying `reverseComplement` to its *forward*
strand sequence is almost always a bad idea.
See the *Finding an arbitrary nucleotide pattern in an entire genome* section
of this document for how to find arbitrary patterns in the *reverse* strand
of a chromosome.

The number of bases in this sequence can be retrieved with:

```{r b5}
chrI <- genome$chrI
length(chrI)
```

Some basic stats:

```{r b6}
afI <- alphabetFrequency(chrI)
afI
sum(afI) == length(chrI)
```

Count all *exact* matches of pattern `"ACCCAGGGC"`:

```{r b7}
p1 <- "ACCCAGGGC"
countPattern(p1, chrI)
```

Like most pattern matching functions in `r Biocpkg('Biostrings')`,
the `countPattern` and `matchPattern` functions
support *inexact* matching. One form of inexact matching is to
allow a few mismatching letters per match. Here we allow at most one:

```{r b8}
countPattern(p1, chrI, max.mismatch=1)
```

With the `matchPattern` function, the locations of the matches are
stored in an `XStringViews` object:

```{r b9}
m1 <- matchPattern(p1, chrI, max.mismatch=1)
m1[4:6]
class(m1)
```

The `mismatch` function (new in `r Biocpkg('Biostrings')` 2)
returns the positions of the mismatching letters for each match:

```{r b10}
mismatch(p1, m1[4:6])
```

Note: The `mismatch` method is in fact a particular case
of a (vectorized) *alignment* function where only "replacements"
are allowed. Current implementation is slow but this will be addressed.

It may happen that a match is *out of limits* like in this example:

```{r b11}
p2 <- DNAString("AAGCCTAAGCCTAAGCCTAA")
m2 <- matchPattern(p2, chrI, max.mismatch=2)
m2[1:4]
p2 == m2[1:4]
mismatch(p2, m2[1:4])
```

The list of exact matches and the list of inexact matches
can both be obtained with:

```{r b12, results='hide'}
m2[p2 == m2]
m2[p2 != m2]
```

Note that the length of `m2[p2 == m2]` should be
equal to `countPattern(p2, chrI, max.mismatch=0)`.

# Finding an arbitrary nucleotide pattern in an entire genome

Now we want to extend our analysis to the *forward* and *reverse*
strands of all the C. elegans chromosomes.
More precisely, here is the analysis we want to perform:

- **The input dictionary:** Our input is a dictionary of 50 patterns. Each
pattern is a short nucleotide sequence of 15 to 25 bases (As, Cs, Gs and Ts
only, no Ns). It is stored in a FASTA file called `"ce2dict0.fa"`. See the
*Finding all the patterns of a constant width dictionary in an entire genome*
section of this document for a very efficient way to deal with the special case
where all the patterns in the input dictionary have the same length.

- **The target:** Our target (or subject) is the *forward* and *reverse* strands
of the seven C. elegans chromosomes (14 sequences in total). We want to find and
report all occurences (or hits) of every pattern in the target. Note that a
given pattern can have 0, 1 or several hits in 0, 1 or 2 strands of 0, 1 or
several chromosomes.

- **Exact or inexact matching?** We are interested in exact matches only (for now).

- **The output:** We want to put the results of this analysis in a file so we
can send it to our collaborators for some post analysis work. Our collaborators
are not necessarily familiar with R or Bioconductor so dumping a high-level R
object (like a list or a data frame) into an .rda file is not an option. For
maximum portability (one of our collaborators wants to use Microsoft Excel for
the post analysis) we choose to put our results in a tabulated file where one
line describes one hit. The columns (or fields) of this file will be (in this
order):
  - **seqname:** the name of the chromosome where the hit occurs.
  - **start:** an integer giving the starting position of the hit.
  - **end:** an integer giving the ending position of the hit.
  - **strand:** a plus (`+`) for a hit in the positive strand or a minus (`-`)
  for a hit in the negative strand.
  - **patternID:** we use the unique ID provided for every pattern in the
  `"ce2dict0.fa"` file.

Let's start by loading the input dictionary with:

```{r c1}
ce2dict0_file <- system.file("extdata", "ce2dict0.fa", package="BSgenome")
ce2dict0 <- readDNAStringSet(ce2dict0_file, "fasta")
ce2dict0
```

Here is how we can write the functions that will perform our analysis:

```{r c2}
writeHits <- function(seqname, matches, strand, file="", append=FALSE)
{
    if (file.exists(file) && !append)
        warning(
            "existing file ", file, " will be overwritten with 'append=FALSE'"
        )
    if (!file.exists(file) && append)
        warning("new file ", file, " will have no header with 'append=TRUE'")
    hits <- data.frame(seqname=rep.int(seqname, length(matches)),
                       start=start(matches),
                       end=end(matches),
                       strand=rep.int(strand, length(matches)),
                       patternID=names(matches),
                       check.names=FALSE)
    write.table(hits, file=file, append=append, quote=FALSE, sep="\t",
                row.names=FALSE, col.names=!append)
}

runAnalysis1 <- function(dict0, outfile="")
{
    library(BSgenome.Celegans.UCSC.ce2)
    genome <- BSgenome.Celegans.UCSC.ce2
    seqnames <- seqnames(genome)
    seqnames_in1string <- paste(seqnames, collapse=", ")
    cat("Target:", metadata(genome)$genome,
        "chromosomes", seqnames_in1string, "\n")
    append <- FALSE
    for (seqname in seqnames) {
        subject <- genome[[seqname]]
        cat(">>> Finding all hits in chromosome", seqname, "...\n")
        for (i in seq_len(length(dict0))) {
            patternID <- names(dict0)[i]
            pattern <- dict0[[i]]
            plus_matches <- matchPattern(pattern, subject)
            names(plus_matches) <- rep.int(patternID, length(plus_matches))
            writeHits(seqname, plus_matches, "+", file=outfile, append=append)
            append <- TRUE
            rcpattern <- reverseComplement(pattern)
            minus_matches <- matchPattern(rcpattern, subject)
            names(minus_matches) <- rep.int(patternID, length(minus_matches))
            writeHits(seqname, minus_matches, "-", file=outfile, append=append)
        }
        cat(">>> DONE\n")
    }
}
```

Some important notes about the implementation of the `runAnalysis1` function:

- `subject <- genome[[seqname]]` is the code that actually loads a chromosome
sequence into memory. Using only one sequence at a time is a good practice to
avoid memory allocation problems on a machine with a limited amount of memory.
For example, loading all the human chromosome sequences in memory would require
more than 3GB of memory!

- We have 2 nested `for` loops: the outer loop walks thru the target (7
chromosomes) and the inner loop walks thru the set of patterns. Doing the other
way around would be very inefficient, especially with a bigger number of
patterns because this would require to load each chromosome sequence into memory
as many times as the number of patterns. `runAnalysis1` loads each sequence only
once.

- We find the matches in the minus strand (`minus_matches`) by first taking the
reverse complement of the current pattern (with
`rcpattern <- reverseComplement(pattern)`) and NOT by taking the reverse
complement of the current subject.

Now we are ready to run the analysis and put the results in the
`"ce2dict0_ana1.txt"` file:

```{r c3}
ofile <- file.path(tempdir(), "ce2dict0_ana1.txt")
runAnalysis1(ce2dict0, outfile = ofile)
```

Here is some very simple example of post analysis:

Get the total number of hits:

```{r c4}
hits1 <- read.table(ofile, header=TRUE)
nrow(hits1)
```

Get the number of hits per chromosome:

```{r c5}
table(hits1$seqname)
```

Get the number of hits per pattern:

```{r c6}
hits1_table <- table(hits1$patternID)
hits1_table
```

Get the pattern(s) with the higher number of hits:

```{r c7}
hits1_table[hits1_table == max(hits1_table)] # pattern(s) with more hits
```

Get the pattern(s) with no hits:

```{r c8}
setdiff(names(ce2dict0), hits1$patternID) # pattern(s) with no hits
```

And finally a function that can be used to plot the hits:

```{r c9}
plotGenomeHits <- function(bsgenome, seqnames, hits)
{
    chrlengths <- seqlengths(bsgenome)[seqnames]
    XMAX <- max(chrlengths)
    YMAX <- length(seqnames)
    plot.new()
    plot.window(c(1, XMAX), c(0, YMAX))
    axis(1)
    axis(
        2,
        at=seq_len(length(seqnames)),
        labels=rev(seqnames),
        tick=FALSE,
        las=1
    )
    ## Plot the chromosomes
    for (i in seq_len(length(seqnames)))
        lines(c(1, chrlengths[i]), c(YMAX + 1 - i, YMAX + 1 - i), type="l")
    ## Plot the hits
    for (i in seq_len(nrow(hits))) {
        seqname <- hits$seqname[i]
        y0 <- YMAX + 1 - match(seqname, seqnames)
        if (hits$strand[i] == "+") {
            y <- y0 + 0.05
            col <- "red"
        } else {
            y <- y0 - 0.05
            col <- "blue"
        }
        lines(c(hits$start[i], hits$end[i]), c(y, y), type="l", col=col, lwd=3)
    }
}
```

Plot the hits found by `runAnalysis1` with:

```{r c10, eval=FALSE}
plotGenomeHits(genome, seqnames(genome), hits1)
```

# Some precautions when using `matchPattern`

Improper use of `matchPattern` (or `countPattern`) can affect performance.

If needed, the `matchPattern` and `countPattern` methods convert their first
argument (the pattern) to an object of the same class than their second argument
(the subject) before they pass it to the subroutine that actually implements the
fast search algorithm.

So if you need to reuse the same pattern a high number of times, it's a good
idea to convert it *before* to pass it to the `matchPattern` or `countPattern`
method. This way the conversion is done only once:

```{r d1}
library(hgu95av2probe)
tmpseq <- DNAStringSet(hgu95av2probe$sequence)
someStats <- function(v)
{
    GC <- DNAString("GC")
    CG <- DNAString("CG")
    sapply(seq_len(length(v)),
           function(i) {
               y <- v[[i]]
               c(alphabetFrequency(y)[1:4],
                 GC=countPattern(GC, y),
                 CG=countPattern(CG, y))
           }
    )
}
someStats(tmpseq[1:10])
```

# Masking the chromosome sequences

Starting with Bioconductor 2.2, the chromosome sequences in a *BSgenome data
package* can have built-in masks.

Starting with Bioconductor 2.3, there can be up to 4 built-in masks per
sequence. These will always be (in this order): (1) the mask of assembly gaps,
(2) the mask of intra-contig ambiguities, (3) the mask of repeat regions that
were determined by the RepeatMasker software, and (4) the mask of repeat regions
that were determined by the Tandem Repeats Finder software (where only repeats
with period less than or equal to 12 were kept).

For a given package, all the sequences will always have the same number of
masks.

```{r f1}
library(BSgenome.Hsapiens.UCSC.hg38.masked)
genome <- BSgenome.Hsapiens.UCSC.hg38.masked
chrY <- genome$chrY
chrY
chrM <- genome$chrM
chrM
```

The built-in masks are named consistenly across all the BSgenome data packages
available in Bioconductor:

| Name  | Active by default | Short description | Long description |
|-------|-------------------|-------------------|------------------|
| AGAPS | yes | assembly gaps | Masks the big N-blocks that have been placed between the contigs during the assembly. This mask is consistent with the Gap track from UCSC Genome Browser. |
| AMB   | yes | intra-contig ambiguities | Masks any IUPAC ambiguity letter that was found in the contig regions of the original sequence. Note that only As, Cs, Gs and Ts remain unmasked when the AGAPS and AMB masks are both active (before SNPs are eventually injected, see below). |
| RM    | no | RepeatMasker | Masks the repeat regions determined by the RepeatMasker software. This mask is consistent with the RepeatMasker track from UCSC Genome Browser. |
| TRF   | no | Tandem Repeats Finder | Masks the tandem repeat regions that were determined by the Tandem Repeats Finder software (with period of 12 or less). |

Table: The built-in masks provided by the BSgenome data packages.

When displaying a masked sequence (here a `MaskedDNAString` object), the *masked
width* and *masked ratio* are reported for each individual mask, as well as for
all the masks together, and for all the active masks together.
The *masked width* is the total number of nucleotide positions that are masked
and the *masked ratio* is the *masked width* divided by the length of the
sequence.

To activate a mask, use the `active` replacement method in conjunction with the
`masks` method. For example, to activate the RepeatMasker mask, do:

```{r f2}
active(masks(chrY))["RM"] <- TRUE
chrY
```

As you can see, the *masked width* for all the active masks together (i.e. the
total number of nucleotide positions that are masked by at least one active
mask) is now the same as for the first mask. This represents a *masked ratio* of
about 83%.

Now when we use a function that is *mask aware*, like `alphabetFrequency`, the
masked regions of the input sequence are ignored:

```{r f3}
active(masks(chrY)) <- FALSE
active(masks(chrY))["AGAPS"] <- TRUE
alphabetFrequency(unmasked(chrY))
alphabetFrequency(chrY)
```

This output indicates that, for this chromosome, the assembly gaps correspond
exactly to the regions in the sequence that were filled with the letter N. Note
that this is not always the case: sometimes Ns, and other IUPAC ambiguity
letters, can be found inside the contigs.

When coercing a `MaskedXString` object to an `XStringViews` object, each
non-masked region in the original sequence is converted into a view on the
sequence:

```{r f4}
as(chrY, "XStringViews")
```

This can be used in conjonction with the `gaps` method to see the gaps between
the views i.e. the masked regions themselves:

```{r f5, results='hide'}
gaps(as(chrY, "XStringViews"))
```

To extract the sizes of the assembly gaps:

```{r f6}
width(gaps(as(chrY, "XStringViews")))
```

Note that, if applied directly to `chrY`, `gaps` returns a `MaskedDNAString`
object with a single mask masking the regions that are not masked in the
original object:

```{r f7}
gaps(chrY)
alphabetFrequency(gaps(chrY))
```

In fact, for any `MaskedDNAString` object, the following should always be
`TRUE`, whatever the masks are:

```{r f8}
af0 <- alphabetFrequency(unmasked(chrY))
af1 <- alphabetFrequency(chrY)
af2 <- alphabetFrequency(gaps(chrY))
all(af0 == af1 + af2)
```

With all chrY masks active:

```{r f9}
active(masks(chrY)) <- TRUE
af1 <- alphabetFrequency(chrY)
af1
gaps(chrY)
af2 <- alphabetFrequency(gaps(chrY))
af2
all(af0 == af1 + af2)
```

Now let's compare three different ways of finding all the occurences of the
`"CANNTG"` consensus sequence in chrY. The Ns in this pattern need to be treated
as wildcards i.e. they must match any letter in the subject.

Without the mask feature, the first way to do it would be to use the
`fixed=FALSE` option in the call to `matchPattern` (or `countPattern`):

```{r f10}
Ebox <- "CANNTG"
active(masks(chrY)) <- FALSE
countPattern(Ebox, chrY, fixed=FALSE)
```

The problem with this method is that the Ns in the subject are also treated as
wildcards hence the abnormally high number of matches. A better method is to
specify the *side* of the matching problem (i.e. *pattern* or *subject*) where
the Ns should be treated as wildcards:

```{r f11}
countPattern(Ebox, chrY, fixed=c(pattern=FALSE,subject=TRUE))
```

Finally, `countPattern` being *mask aware*, this can be achieved more
efficiently by just masking the assembly gaps and ambiguities:

```{r f12}
active(masks(chrY))[c("AGAPS", "AMB")] <- TRUE
alphabetFrequency(chrY, baseOnly=TRUE)  # no ambiguities
countPattern(Ebox, chrY, fixed=FALSE)
```

Note that some chromosomes can have Ns outside the assembly gaps:

```{r f13}
chr2 <- genome$chr2
active(masks(chr2))[-2] <- FALSE
alphabetFrequency(gaps(chr2))
```

so it is recommended to always keep the AMB mask active (in addition to the
AGAPS mask) whatever the sequence is.

Note that not all functions that work with an `XString` input are *mask aware*
but more will be added in the near future. However, most of the times there is a
alternate way to exclude some arbitrary regions from an analysis without having
to use *mask aware* functions. This is described below in the *Hard masking*
section.

# Hard masking

coming soon...

# Injecting known SNPs in the chromosome sequences

coming soon...

# Finding all the patterns of a constant width dictionary in an entire genome

The `matchPDict` function can be used instead of `matchPattern` for the kind of
analysis described in the
*Finding an arbitrary nucleotide pattern in an entire genome* section but it
will be much faster (between 100x and 10000x faster depending on the size of the
input dictionary).

Note that a current limitation of `matchPDict` is that it only works with a
dictionary of DNA patterns where all the patterns have the same number of
nucleotides (constant width dictionary).

See `?matchPDict` for more information.

Here is how our `runAnalysis1` function can be modified in order to use
`matchPDict` instead of `matchPattern`:

```{r e1}
runOneStrandAnalysis <- function(dict0, bsgenome, seqnames, strand,
                                 outfile="", append=FALSE)
{
    cat("\nTarget: strand", strand, "of", metadata(bsgenome)$genome,
        "chromosomes", paste(seqnames, collapse=", "), "\n")
    if (strand == "-")
        dict0 <- reverseComplement(dict0)
    pdict <- PDict(dict0)
    for (seqname in seqnames) {
        subject <- bsgenome[[seqname]]
        cat(
            ">>> Finding all hits in strand",
            strand,
            "of chromosome",
            seqname,
            "...\n"
        )
        mindex <- matchPDict(pdict, subject)
        matches <- extractAllMatches(subject, mindex)
        writeHits(seqname, matches, strand, file=outfile, append=append)
        append <- TRUE
        cat(">>> DONE\n")
    }
}

runAnalysis2 <- function(dict0, outfile="")
{
    library(BSgenome.Celegans.UCSC.ce2)
    genome <- BSgenome.Celegans.UCSC.ce2
    seqnames <- seqnames(genome)
    runOneStrandAnalysis(
        dict0, genome, seqnames, "+", outfile=outfile, append=FALSE
    )
    runOneStrandAnalysis(
        dict0, genome, seqnames, "-", outfile=outfile, append=TRUE
    )
}
```

Remember that `matchPDict` only works if all the patterns in the input
dictionary have the same length so for this 2nd analysis, we will truncate the
patterns in `ce2dict0` to 15 nucleotides:

```{r e2}
ce2dict0cw15 <- DNAStringSet(ce2dict0, end=15)
```

Now we can run this 2nd analysis and put the results in the
`"ce2dict0cw15_ana2.txt"` file:

```{r e3}
ofile <- file.path(tempdir(), "ce2dict0cw15_ana2.txt")
runAnalysis2(ce2dict0cw15, outfile = ofile)
```

# Session info

```{r g1}
sessionInfo()
```
